---
title: "p8105_hw3_hl3859"
author: "Hanrui Li"
date: "2024-10-16"
output: github_document
---

# Problem 1

### Import the dataset

```{r message=FALSE}
library(tidyverse)
library(p8105.datasets)

data("ny_noaa")
```

The `ny_noaa` has `r nrow(ny_noaa)` observations and `r ncol(ny_noaa)` variables.

-   `date`: Observation date (e.g., `r ny_noaa$date[1]`)

-   `prcp`: Precipitation (tenths of mm)

-   `snow`: Snowfall (mm)

-   `snwd`: Snow depth (mm)

-   `tmax`: Maximum temperature (tenths of degree Celsius)

-   `tmin`: Maximum and minimum temperature (tenths of degree Celsius)

There are `r sum(is.na(ny_noaa))` missing values.

### Clean the dataset

```{r}
ny_noaa_cleaned <- ny_noaa |>
  mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date),
    tmax = as.numeric(tmax) / 10,
    tmin = as.numeric(tmin) / 10,
    prcp = as.numeric(prcp) / 10
  )

ny_noaa_cleaned |>
  count(snow) |>
  drop_na(snow) |>
  arrange(desc(n)) |>
  head(10) |>
  knitr::kable()
```

The most common values for snowfall typically include 0 (2008508 observations), 25 (31022 observations), and 13 (23095 observations), as most days may not experience snow, particularly in non-winter months or in locations where snowfall is rare.

### Show the average max temperature in January and in July in each station across years

```{r message=FALSE, warning=FALSE}
ny_noaa_cleaned |>
  filter(month %in% c(1, 7)) |>
  group_by(year, month, id) |>
  summarize(avg_tmax = mean(tmax, na.rm = TRUE), .groups = "drop") |>
  mutate(month = factor(month, labels = c("January", "July"))) |>
  ggplot(aes(x = year, y = avg_tmax, color = factor(year))) +
  geom_boxplot() +
  facet_grid(.~ month) +
  labs(
    title = "Average Max Temperature in January and July across Stations",
    x = "Year",
    y = "Average Max Temperature (°C)"
  ) +
  theme(legend.position = "none")
```

January shows lower average temperatures (often below freezing) and July shows higher average temperatures (mostly above 20°C). This is reasonable as January is in winter and July is in summer.

There are some outliers (shown as dots) in the January panel, where the maximum temperatures drop significantly below 0°C. In July, there are notably lower maximum temperatures, likely representing cool summers.

### (i) `tmax` vs `tmin` for the full dataset

```{r message=FALSE, warning=FALSE}
ggplot(ny_noaa_cleaned, aes(x = tmin, y = tmax)) +
  geom_bin2d() + 
  scale_fill_viridis_c() +
  labs(
    title = "Relationship Between Max and Min Temperatures",
    x = "Minimum Temperature (°C)",
    y = "Maximum Temperature (°C)"
  )
```

### (ii) distribution of snowfall values greater than 0 and less than 100 separately by year

```{r}
ny_noaa_cleaned |>
  filter(snow > 0, snow < 100) |>
  group_by(year) |>
  ggplot(aes(x = factor(year), y = snow)) +
  geom_violin() +
  labs(
    title = "Distribution of Snowfall (0 < Snow < 100 mm) by Year",
    x = "Year",
    y = "Snowfall (mm)"
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# Problem 2

### Load, tidy, merge, and otherwise organize the data sets

```{r}
library(tidyverse)
library(ggplot2)

demographic <- read.csv("data/nhanes_covar.csv", na = c("NA", ",", ""), skip = 4) |>
  janitor::clean_names() |>
  mutate(sex = factor(case_when(sex == 1 ~ "male", sex == 2 ~ "female")),
         education = factor(case_when(education == 1 ~ "Less than high school", 
                                      education == 2 ~ "High school equivalent", 
                                      education == 3 ~ "More than high school")))

accel <- read.csv("data/nhanes_accel.csv", na = c("NA", ",", "")) |>
  janitor::clean_names()

demographic_cleaned <- demographic |>
  filter(age >= 21) |>
  drop_na()

merged <- merge(demographic_cleaned, accel, by = "seqn")
```

### Produce a reader-friendly table for the number of men and women in each education category

```{r message=FALSE}
merged |>
  group_by(education, sex) |>
  janitor::tabyl(education, sex) |>
  knitr::kable()
```

For Less than high school, the number of participants is nearly balanced between males (27) and females (28). For High school equivalent, there is a slight gender imbalance with more males (35) than females (23). For More than high school, both males (56) and females (59) are well represented, showing a fairly even distribution across genders.

In general, participants with "More than high school" education make up the largest group, followed by "High school equivalent," with "Less than high school" having the fewest participants. This reflects the general trend of higher educational attainment among the participants.

### Create a visualization of the age distributions for men and women in each education category

```{r}
ggplot(merged, aes(x = age, fill = sex)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ education) +
  labs(title = "Age Distribution by Sex and Education Level",
       x = "Age",
       y = "Density") +
  theme_minimal()
```

For High school equivalent, males and females show a similar age distribution, with a large concentration of participants between 40 and 60 years. There is a small divergence in older ages, with slightly more males in their early 40s and late 50s.

For Less than high school, the age distribution skews older for both males and females. The peak for females is around 70 years, while for males, it is slightly lower, around 60. This could indicate that older participants tend to have lower education levels.

For More than high school, participants are more evenly distributed across age groups for both genders, with more participants in their 50s and 60s. Females tend to dominate the older age group (70+), while males are more evenly spread across age ranges.

Males tend to have a slightly wider spread across age groups in all education levels, while females tend to peak more sharply in specific age ranges, particularly among the less educated. As education level increases, the age distribution becomes broader, indicating that participants with higher education levels are more evenly distributed across various age groups. Conversely, those with lower education levels tend to be older. This may suggest that older generations had less access to education compared to younger participants.

### Focus on total activity

```{r message=FALSE}
merged <- merged |>
  mutate(total_activity = rowSums(select(merged, starts_with("min"))))

ggplot(merged, aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.6) + 
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ education) + 
  labs(title = "Total Activity vs Age by Sex and Education Level",
       x = "Age",
       y = "Total Activity (MIMS)") +
  theme_minimal() +
  theme(legend.position = "right")
```

Across all education levels, total activity tends to decline with age. This is most noticeable among participants older than 60, where both males and females show a sharp drop in total activity. For younger participants (around 20–40 years), activity levels appear more varied, particularly in the "High school equivalent" and "More than high school" groups. This could indicate a broader range of physical activity habits in younger individuals.

For High School Equivalent, males and females have similar levels of activity, with a slight decline in total activity for both genders as they approach middle age. However, there are spikes in activity around ages 40-50 for females and a more consistent decline for males.

For Less than High School, females tend to have higher activity levels than males from age 20 to 50, but this changes after age 50, where both males and females show a steep drop-off in activity.

For More than High School, there is a slight fluctuation in activity for both males and females around middle age, but total activity tends to plateau or decline slightly after the age of 50. This group shows the most consistency in activity levels for both genders, with less variation in total activity across age groups. This suggests that participants with higher education levels may have more stable physical activity patterns.

In all groups, there are notable outliers, particularly in younger age ranges, where some individuals have very high total activity levels compared to others.

### Make a three-panel plot that shows the 24-hour activity time courses for each education level

```{r}
merged |>
  pivot_longer(cols = starts_with("min"),
               names_to = "Time",
               values_to = "Activity") |>
  ggplot(aes(x = as.numeric(gsub("min", "", Time)), y = Activity, color = sex)) +
  geom_line(stat = "summary", fun = "mean") +
  facet_wrap(~ education) +
  labs(title = "24-hour Activity Time Courses by Education Level",
       x = "Time of Day (minutes from midnight)",
       y = "Average Activity (MIMS)") +
  theme_minimal()
```

The participants, regardless of education level, tend to follow a daily pattern of higher activity in the morning and afternoon, with a drop during the night and a midday lull. This is consistent with typical daily schedules, work hours, and rest periods.

Participants with higher education ("More than high school") have more consistent activity levels across the day, with smaller differences between men and women. This may reflect more regulated daily routines, possibly due to structured work environments or more regular exercise patterns.

Those with "Less than high school" education tend to exhibit higher overall activity during peak times, which could indicate physically demanding jobs or less sedentary lifestyles. Women in this group, in particular, show more pronounced activity peaks, suggesting higher daily physical exertion.

# Problem 3

### Import, clean, and tidy the datasets

```{r message=FALSE}
library(tidyverse)

jan_2020 <- read.csv("data/citibike/Jan 2020 Citi.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  mutate(month = "January", year = 2020)

jan_2024 <- read.csv("data/citibike/Jan 2024 Citi.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  mutate(month = "January", year = 2024)

july_2020 <- read.csv("data/citibike/July 2020 Citi.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  mutate(month = "July", year = 2020)

july_2024 <- read.csv("data/citibike/July 2024 Citi.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  mutate(month = "July", year = 2024)

citi_bike <- bind_rows(jan_2020, jan_2024, july_2020, july_2024) |>
  filter(duration <= 240)
```

The `citi_bike` data has `r nrow(citi_bike)` observations and `r ncol(citi_bike)` variables.

- ride_id: Unique identifier for each ride (e.g., `r citi_bike$ride_id[1]`)

- rideable_type: Type of bike used for the ride, typically either "classic" or "electric"

- weekdays: The day of the week the ride started

- duration: The total duration of the ride in minutes

- start_station_name: The name of the station where the ride started

- end_station_name: The name of the station where the ride ended

- member_casual: Rider type, indicating whether the rider is a Citi Bike member or a casual user

- month: The month in which the ride occurred

- year: The year when the ride took place

### Produce a reader-friendly table showing the total number of rides in each combination of year and month separating casual riders and Citi Bike members

```{r}
citi_bike |>
  group_by(year, month, member_casual) |>
  summarize(total_rides = n()) |>
  knitr::kable()
```

Members are the dominant users of the system, and their ridership has grown significantly from 2020 to 2024. There are more rides in July compared to January, consistent with better summer weather. While members remain the primary users, casual rider usage has also increased over time, indicating broader adoption of the Citi Bike system.

### Make a table of 5 most popular starting stations in July 2024

```{r}
citi_bike |>
  filter(month == "July", year == 2024) |>
  group_by(start_station_name) |>
  summarize(total_rides = n()) |>
  arrange(desc(total_rides)) |>
  head(5) |>
  knitr::kable()
```

### Make a plot of the median ride duration by day of week, month, and year

```{r}
citi_bike |> 
  group_by(year, month, weekdays) |> 
  summarize(median_duration = median(duration), .groups = "drop") |>
  ggplot(aes(x = weekdays, y = median_duration, color = factor(year))) +
  geom_line(aes(group = interaction(year, month)), size = 1) +
  facet_wrap(~ month) +
  labs(title = "Median Ride Duration by Day of the Week, Month, and Year",
       x = "Day of the Week", 
       y = "Median Ride Duration (minutes)") +
  theme_minimal()
```

July has longer ride durations than January, driven by better weather conditions in the summer. Weekend rides (Saturday and Sunday) tend to be longer, particularly in July, likely due to recreational use. Ride durations are generally shorter in 2024 compared to 2020, potentially indicating a shift in rider behavior or an increase in the usage of electric bikes for shorter trips.

### Distribution of ride duration for 2024 by membership status, bike type, and month

```{r}
citi_bike |>
  filter(year == 2024) |>
  ggplot(aes(x = duration, fill = member_casual)) +
  geom_density(alpha = 0.5) +
  facet_grid(rideable_type ~ month) +
  labs(title = "Ride Duration Distribution by Membership Status, Bike Type, and Month in 2024",
       x = "Ride Duration (minutes)", 
       y = "Density")
```

Short trips dominate the Citi Bike system, with most rides being under 10 minutes regardless of the bike type or rider's membership status. Casual riders tend to have slightly longer ride durations than members, possibly due to different use cases. Electric bikes show a similar ride duration distribution for both casual and member riders, likely due to their appeal for short, efficient trips. Seasonal differences are present but subtle, with July showing slightly longer rides, especially for casual riders.

